Relation of severe COVID-19 in Scotland to transmission-related factors and risk conditions eligible for shielding support: REACT-SCOT case-control study

### 2021-08-29 transmissionBMED-D-21-00640.Rmd
reinfections

@@

# Abstract  
* _Background_ -  Clinically vulnerable individuals have been advised to shield themselves during the COVID-19 epidemic.  The objectives of this study were to investigate: (1) the risk of severe COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.  
* _Methods_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.  This dataset (REACT-SCOT) was  linked to the list of `r nrow(shielded.full)` individuals identified as eligible for shielding.  Severe COVID-19 was defined as cases that entered critical care or were fatal.  
* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estci(table.listedgr3.severecases.nocare["Moderate risk condition", 3])` in those with moderate risk conditions and 
`r format.estci(table.listedgr3.severecases.nocare["Eligible for shielding", 3])` 
in those eligible for shielding.  The highest rate was in solid organ transplant recipients: rate ratio `r format.estci(table.shielded.severecases.nocare["Solid organ transplant", 3])`.  Risk of severe COVID-19 increased with the number of adults but decreased with the number of school-age children in the household. Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estci(table.shielded.severecases.nocare["Recent hospital visit/stay", 3])`  overall.   To test for causality, a case-crossover analysis was undertaken; with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estci(table.tw["Recent.TW.only", 4])`. The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.  
* _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.  For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts. Mitigating the impact of the epidemic requires control of nosocomial transmission.

@@

# Background
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.  Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].  Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters advising strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.

@@

Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.  Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome. This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.

@@

We used the Scottish Morbidity Records SMR01 (inpatients and day cases) and SMR00 (outpatient attendance) together with the RAPID database to derive variables encoding recent exposure to hospitals. We defined the variable "recent hospital exposure" as any hospital in-patient stay, day case attendance or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.  Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].  The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.  <!---Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated. --->

@@

To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.  The population attributable risk fraction (PARF) of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.  where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].  <!---Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.--->

@@

# Results
## Shielding eligibility and risk of COVID-19
Supplementary Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.  To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.  Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.

@@

We examined the time course of the rate ratio associated with eligibility for shielding.  As shown in Supplementary Table \ref{tab:letterdates} most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of  shielding letters sent on 3 April 2020.  Supplementary Figure \ref{fig:plotrateratio}(a) shows  the risk ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.  The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding. Thus there was no evidence that shielding advice reduced the rate ratio.

@@

In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r format.estcipv(table.rapid.coeffs["rapid.recent", 3], table.rapid.coeffs["rapid.recent", 4])`]; the  rate ratios associated with day case exposure [`r format.estcipv(table.daycase.coeffs["daycase.recent", 3], table.daycase.coeffs["daycase.recent", 4])`] or outpatient exposure  [`r format.estcipv(table.opd.coeffs["opd.recent", 3], table.opd.coeffs["opd.recent", 4])`]  were  much lower.   To test if the association of severe COVID-19 with recent inpatient exposure was likely to simply reflect  confounding by time invariant comorbidity, the association was examined by time window as described in the Methods section. Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.   
<!--- tabulate underlying cause of death on death certs of fatal cases
paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(hosp.recent, covid_ucod)))
with(cc.severe[CASE==1], table(hosp.recent, inhosp))
paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(occup, covid_ucod)))
--->
The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.  From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.  As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.  The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.

@@

The high frequency of recent hospital exposure in cases of severe COVID-19 who were eligible for shielding, together with the increase in the rate ratio associated with recent hospital  exposure during periods when population-wide social distancing measures were being imposed, suggests that exposure to transmission in hospital settings is at least part of the explanation for the failure to reduce the rate ratio associated with eligibility for shielding.   Our finding that even in those advised to shield the risk of severe disease was increased in those who were sharing a household with other adults suggests that this was another constraint on the effectiveness of advice to shield, as no support was provided for other household members to co-isolate with the vulnerable individual.

### Association with household composition
The increased risk of severe COVID-19 in patient facing  health care workers is consistent with our previous report of a threefold risk for hospitalised COVID-19 earlier in the epidemic [@shah_risk_2020].  We and others have previously reported  lower risk in teachers compared with others of the same age and sex [@mcallister_report_2020; @ludvigsson_open_2021].  The inverse association of severe disease with the number of school-age children in the household extends and confirms the findings of an earlier study of health care workers and their families [@wood_living_2021]. In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 after adjusting for covariates, but no dose-response relationship was reported [@forbes_association_2020].  The inverse association of severe COVID-19 with past exposure to children is consistent with evidence that  other coronaviruses generate cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020]. From a public health perspective the most relevant implication is that although the rate ratio per child has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows. The association of severe COVID-19 with number of adults in the household is consistent with classic studies of other viral infections showing that secondary cases in a household, where the infecting dose is likely to be high, are more severe than index cases [@aaby_malnutrition_1988].

### Association with recent exposure to hospital
A striking finding from our analysis is the  PARF  of  `r round(100 * pc.all * (r.all - 1) / r.all)` % for severe disease associated with recent exposure to hospital.  Although we had pre-specified this category to include day case and outpatient exposure, the association was driven by inpatient exposure.  The PARF associated with inpatient exposure was  `r round(100 * parf.rapid)`%.  Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF.  As we have shown most of this difference is due to the fact that the international definition does not capture cases diagnosed shortly after discharge that could have been acquired in hospital.   The other difference  is that the  international definition classifies  those with onset 5-6 days after admission as indeterminate.  As the average incubation period for COVID-19 is 5-6 days, exposure in this time window is relevant.

The role of nosocomial transmission was recognized early in the epidemic. Hospital-acquired infection was suspected in 41% of hospitalised cases seen in January 2020 in one centre in Wuhan [@wang_clinical_2020-2].  A commentary in April 2020 noted that "in Lombardy, SARS-CoV-2 became largely a nosocomial infection" [@boccia_what_2020].  Most previous reports have focused on the narrow definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].  The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].

A paper presented to SAGE on 28 January 2021 broadened the ECDPC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure" [@knight_contribution_2021]. From linking hospital episode data to COVID-19 tests and CO-CIN, the authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that onward transmission of COVID-19 from hospital-acquired cases could account for another 5% of hospitalised cases.

@@

As all hospital inpatients are tested for SARS-CoV-2 every few days and all deaths within 28 days of a positive test are officially classified as deaths involving COVID-19,  some misclassification of deaths from other causes as COVID-19 deaths is likely in those with recent hospital exposure.  However the rate ratio associated with recent hospital exposure was barely changed when fatal cases without COVID-19 as underlying cause of death on death certificate were omitted.  <!--- Chris R to run a comparison of 28-day fatality rates in those who test positive and those who test negative. --->

@@

```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Rate ratios for severe COVID-19 in those not resident in care homes, by risk group and sliding window of presentation dates.  (b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates. Line thickness is proportional to precision of estimate."}

@@

```{r hcai_hosp, echo=FALSE, warning=FALSE, message=FALSE}
if(FALSE) {
knitr::kable(table.hcai.hosp.recent,
             escape=FALSE, 
             booktabs=TRUE,
             col.names=c("No recent hospital exposure", "Recent hospital exposure"), 
			 label="hcai_hosp",
             row.names=TRUE,
             align=rep("r", 2), 
             caption="Hospital onset status as defined by ECDPC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
     add_footnote(label="Percentages are column percentages",
                 notation="none") %>%
    kable_styling(latex_options=c("hold_position"))
}
```

@@

knitr::kable(table.hcai.rapid.recent,
             escape=FALSE, 
             booktabs=TRUE,
			 label="hcai_rapid",
             row.names=TRUE,
             align=rep("r", 2), 
             caption="Severe test-positive cases classified by hospital onset status (as defined by ECDPC) and recent inpatient exposure") %>%
	 column_spec(column=1, width="4cm") %>%
	 column_spec(column=2, width="2.5cm") %>%
	 column_spec(column=3, width="2.5cm") %>%
#	 column_spec(column=4, width="2.5cm") %>%
#	 column_spec(column=5, width="2.5cm") %>%
     add_footnote(label="Percentages are column percentages",
                 notation="none") %>%
    kable_styling(latex_options=c("HOLD_position"))

>>>>

table.testpos <- paste.rowpercent(with(cc.severe[CASE==1], 
table(SPECIMENDATE > as.Date("2020-08-31"), testpositive.case)))

@@

# Abstract  
* _Background_ -  Clinically vulnerable individuals have been advised to shield themselves during the COVID-19 epidemic.  The objectives of this study were to investigate: (1) the risk of severe COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.  
* _Methods_ -  In a matched case-control design, all `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.  This dataset (REACT-SCOT) was  linked to the list of `r nrow(shielded.full)` individuals identified as eligible for shielding.  Severe COVID-19 was defined as cases that entered critical care or were fatal. Rate ratios were estimated by conditional logistic regression.  
* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estci(table.listedgr3.severecases.nocare["Moderate risk condition", 3])` in those with moderate risk conditions and 
`r format.estci(table.listedgr3.severecases.nocare["Eligible for shielding", 3])` 
in those eligible for shielding.  The highest rate was in solid organ transplant recipients: rate ratio `r format.estci(table.shielded.severecases.nocare["Solid organ transplant", 3])`.  Risk of severe COVID-19 increased with the number of adults but decreased with the number of school-age children in the household. Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estci(table.shielded.severecases.nocare["Recent hospital visit/stay", 3])`  overall.   The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.  
* _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household. Mitigating the impact of the epidemic requires control of nosocomial transmission.

@@

# Background
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.  Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].  Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters advising strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots. Compliance was voluntary.

@@

Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome. This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing. In the first wave (up end of August 2020), `r table.testpos[1, 2]` of the severe cases were test-positive; in the second wave beginning on 1 September 2020, `r table.testpos[2, 2]` were test-positive.

Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.

@@

We used the Scottish Morbidity Records SMR01 (inpatients and day cases) and SMR00 (outpatient attendance) together with the RAPID database to derive variables encoding recent exposure to hospitals. We defined the variable "recent hospital exposure" as any hospital in-patient stay, day case attendance or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired. In Scotland people with COVID-19 symptoms were instructed not to visit hospitals, to attend designated test centres for testing within 3 days, and to call an emergency number for admission if their symptoms worsened.  It is therefore unlikely that hospital visits by people with COVID-19 symptoms could account for the association of first positive tests 5-6 days before.

Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].  The ECDC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.  <!---Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated. --->

@@

To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.  The population attributable risk fraction (PARF) of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.  where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].

@@

# Results
## Shielding eligibility and risk of COVID-19
Supplementary Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.  To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been allocated to the "Additional conditions" category; there were no severe cases in this group.  Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.

@@

We examined the time course of the rate ratio associated with eligibility for shielding.  As shown in Supplementary Table \ref{tab:letterdates} most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of  shielding letters sent on 3 April 2020.  Supplementary Figure \ref{fig:plotrateratio}(a) shows  the rate ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.  The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding. Thus there was no evidence that shielding advice reduced the rate ratio.

@@

In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r format.estcipv(table.rapid.coeffs["rapid.recent", 3], table.rapid.coeffs["rapid.recent", 4])`]; the  rate ratios associated with day case exposure [`r format.estcipv(table.daycase.coeffs["daycase.recent", 3], table.daycase.coeffs["daycase.recent", 4])`] or outpatient exposure  [`r format.estcipv(table.opd.coeffs["opd.recent", 3], table.opd.coeffs["opd.recent", 4])`]  were  much lower.   To test if the association of severe COVID-19 with recent inpatient exposure was likely to simply reflect  confounding by time invariant comorbidity, the association was examined by time window as described in the Methods section. Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.

The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.  From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.  As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.  The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDC classification was `r round(r.hcai, 1)`.

@@

The high frequency of recent hospital exposure in cases of severe COVID-19 who were eligible for shielding, together with the increase in the rate ratio associated with recent hospital  exposure during periods when population-wide social distancing measures were being imposed, suggests that exposure to transmission in hospital settings is at least part of the explanation for the failure to reduce the rate ratio associated with eligibility for shielding.   Our finding that even in those advised to shield the risk of severe disease was increased in those who were sharing a household with other adults suggests that this was another constraint on the effectiveness of advice to shield, as no support was provided for other household members to co-isolate with the vulnerable individual. A survey conducted online in the first two weeks of June 2020 found that 65% of those living alone or sharing a a household with one other person reported that they followed the guidance completely, but only 55% of those living with with two or more others [@delvaux_covid-19_2020].

### Association with household composition
The increased risk of severe COVID-19 in patient facing  health care workers is consistent with our previous report of a threefold risk for hospitalised COVID-19 earlier in the epidemic [@shah_risk_2020].  We and others have previously reported  lower risk in teachers compared with others of the same age and sex [@mcallister_report_2020; @ludvigsson_open_2021].  The inverse association of severe disease with the number of school-age children in the household extends and confirms the findings of an earlier study of health care workers and their families [@wood_sharing_2021]. In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 after adjusting for covariates, but no dose-response relationship was reported [@forbes_association_2020].  The inverse association of severe COVID-19 with past exposure to children is consistent with evidence that  other coronaviruses generate cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020]. From a public health perspective the most relevant implication is that although the rate ratio per child has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows. The association of severe COVID-19 with number of adults in the household is consistent with classic studies of other viral infections showing that secondary cases in a household, where the infecting dose is likely to be high, are more severe than index cases [@aaby_malnutrition_1988].

### Association with recent exposure to hospital
A striking finding from our analysis is the  PARF  of  `r round(100 * pc.all * (r.all - 1) / r.all)`% for severe disease associated with recent exposure to hospital.  Although we had pre-specified this category to include day case and outpatient exposure, the association was driven by inpatient exposure.  The PARF associated with inpatient exposure was  `r round(100 * parf.rapid)`%.  Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF.  As we have shown most of this difference is due to the fact that the international definition does not capture cases diagnosed shortly after discharge that could have been acquired in hospital.   The other difference  is that the  international definition classifies  those with onset 5-6 days after admission as indeterminate.  As the average incubation period for COVID-19 is 5-6 days, exposure in this time window is relevant.

The role of nosocomial transmission was recognized early in the epidemic. Hospital-acquired infection was suspected in 41% of hospitalised cases seen in January 2020 in one centre in Wuhan [@wang_clinical_2020-2].  A commentary in April 2020 noted that "in Lombardy, SARS-CoV-2 became largely a nosocomial infection" [@boccia_what_2020].  Most previous reports have focused on the narrow definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].  The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents) during 2020, 30% (1015/3389) met the  ECDC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].

A paper presented to SAGE on 28 January 2021 broadened the ECDC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure" [@knight_contribution_2021]. From linking hospital episode data to COVID-19 tests and CO-CIN, the authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that onward transmission of COVID-19 from hospital-acquired cases could account for another 5% of hospitalised cases.

@@

As all hospital inpatients are tested for SARS-CoV-2 every few days and all deaths within 28 days of a positive test are officially classified as deaths involving COVID-19,  some misclassification of deaths from other causes as COVID-19 deaths is likely in those with recent hospital exposure.  However the rate ratio associated with recent hospital exposure was barely changed when fatal cases without COVID-19 as underlying cause of death on death certificate were omitted.

@@

```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Rate ratios for severe COVID-19 in those not resident in care homes, by risk group and sliding window of presentation dates.  (b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates. Line thickness is proportional to precision of estimate. Labels on the y axes are untransformed rate ratios, though the axis scales are log-transformed"}

@@

```{r hcai_hosp, echo=FALSE, warning=FALSE, message=FALSE}
if(FALSE) {
knitr::kable(table.hcai.hosp.recent,
             escape=FALSE, 
             booktabs=TRUE,
             col.names=c("No recent hospital exposure", "Recent hospital exposure"), 
			 label="hcai_hosp",
             row.names=TRUE,
             align=rep("r", 2), 
             caption="Hospital onset status as defined by ECDC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
     add_footnote(label="Percentages are column percentages",
                 notation="none") %>%
    kable_styling(latex_options=c("hold_position"))
}
```

@@

knitr::kable(table.hcai.rapid.recent,
             escape=FALSE, 
             booktabs=TRUE,
			 label="hcai_rapid",
             row.names=TRUE,
             align=rep("r", 2), 
             caption="Severe test-positive cases classified by hospital onset status (as defined by ECDC) and recent inpatient exposure") %>%
	 column_spec(column=1, width="4cm") %>%
	 column_spec(column=2, width="2.5cm") %>%
	 column_spec(column=3, width="2.5cm") %>%
    add_footnote(label=c("Percentages are column percentages",
                         "ECDC, European Centre for Disease Prevention and Control"),
                 notation="none") %>%
    kable_styling(latex_options=c("HOLD_position"))
